Data Specialization for a Staged Lambda Calculus

### 2015-01-27 paper/splitting.tex
some work on stage one splitting and general syntax stuff

\begin{concretesyntax}

@@

Splitting a stage~\bbone\ term $e$ in \lang\ yields a term 
$e_1$ that performs all stage~\bbone\ subcomputations in $e$ and a \emph{resumer term} $e_2 = l.r$ that performs all stage~\bbtwo\ subcomputations.
To ensure the results of splitting produce the same output as the evaluation $e\mathbin{\redonesym}[\xi;v]$,
we require that $e_1$ reduce to the tuple $(y,b)$,
where $y$, called the \emph{\bbone-result} of $e$, is identical to $\masko{\xi;v}$, and that
$\letin l b r$ reduces to the same value as $\maskt{\xi;v}$. (We refer to $b$ as the \emph{boundary value} of the split since it carries information between stage~\bbone\ and stage~\bbtwo\ subcomputations.) Since~$e_1$ performs both stage~\bbone\ subcomputations needed to produce the \bbone-result and subcomputations (refered to as \emph{precomputations}) that produce the boundary value, we refer to this term as the \emph{combined term}, and hereafter denote it as~$c$.

The splitting algorithm for stage~\bbone\ terms, as specified by the
judgment $e \splitonesym [c,l.r]$ in \cref{fig:termSplit}, proceeds
recursively on the structure of~$e$.
When $e$ is a terminal (a variable or base constant)
splitting yields a combined term formed by tupling $e$ with a \texttt{()} precomputation, and the trivial resumer \texttt{()}. (Stage~\bbone\ terminals, by definition, contain no stage~\bbtwo\ subcomputations.)  For example, the integer constant \texttt{3} splits into the combined term \texttt{(3,())} and resumer \texttt{\_=>()}.

% Note the direct correspondence of these splitting outputs to the output of the stage~\bbone\ and~\bbtwo\ masking functions.

For all non-terminals (except $\next$),
splitting descends into $e$, recursively splitting its $n$ subterms
to produce their respective combined terms $c_1,\ldots,c_n$ and resumers $r_1, \ldots, r_n$.
The combined term of $e$ is formed by binding $c_1,\ldots,c_n$
to the patterns $(y_1,b_1),\ldots,(y_n,b_n)$
to isolate \bbone-results from boundary values. Then,
the \bbone-result of $e$ is formed by replacing $e$'s subterms with $y_1,\ldots,y_n$.
The resumer binds the boundary values $b_1,\ldots,b_n$ to an
argument $(l_1,\ldots,l_n)$ in a term that has the same structure
of~$e$ but where each subterm is replaced by its resumer ($r_i$'s).

@@

\subsection{Improved Definition}
\label{sec:resumers}

The definitions of masking and splitting can be augmented to better handle special cases,
particularly ones that depend on the type of the term we're splitting.
One such special case is the behavior of stage \bbone\ splitting at stage \bbone\ types that contain no $\fut$s or $\to$s.
We call such types {\em monostage} (judgmentally, $A~mono$), 
because a partial value $v$ with a monostage type can contain no $\next$s,
and so must comprise only stage \bbone\ information.

Intuitively, for $[\xi;v] : A$ and $A~mono$, 
we would expect that $\maskt{\xi;v}$ only reduce to something that carries no information.
This indeed happens: for example, $\maskt {\cdot;\mathtt{(inl~3,5)}}$ reduces to \texttt{((),())}.
But while \texttt{((),())} technically ``carries no information," it still has non-trivial structure.
This is not only unintuitive, but potentially expensive.
Instead, we desire a stronger property, that $\maskt{\xi;v}$ only reduce to exactly \texttt{()}. 
We can achieve this by adjusting just the tuple rule:
\[
\maskt{\cdot;(v_1,v_2)} = \left\{ \begin{array}{ll} 
\mathtt{()} &\text{if } (v_1,v_2):A~mono\\ 
(\maskt{\cdot;v_1},\maskt{\cdot;v_2}) &\text{otherwise} \end{array} \right.
\]
From here, the splitting rules can be made to comport with this new 
definition adding special-case tuple, projection, and application 
rules:
\begin{mathpar}
\infer
	{(e_1,e_2):A\times B\splitonesym [\ldots,(l_1,l_2).(r_1;r_2;())]}
	{e_1 : A \splitonesym [c_1,l_1.r_1] 
	& e_2 : B \splitonesym [c_2,l_2.r_2] 
	& A \times B~mono} 
\and
\infer
	{\pio~e:A\splitonesym [\ldots,l.(r;())]}
	{e : A\times B \splitonesym [c,l.r] 
	& A \times B~mono} 
\and
\infer
	{e_1~e_2: B\splitonesym [\ldots,(l_1,l_2,l_3).(r_1 (r_2,l_3); ())]}
	{e_1 : A \to B \splitonesym [c_1,l_1.r_1] 
	& e_2 : A \splitonesym [c_2,l_2.r_2] 
	& B~mono} 
\end{mathpar}
The old tuple, projection, and application rules 
must also be augmented to apply only when the $mono$ premises don't hold.
Also, all splitting rules need to be augmented to carry types, though
we omit the contexts here.

\end{concretesyntax}

>>>>

\begin{abstrsyn}

@@

Recall that a term $\colone e A$ represents a multistage computation with a multistage result.
To represent this in a separated form, we use the construct $\pipeM c l r$, 
where $c$ is a monostage term and $r$ is a monostage term open on the variable $l$.
This ``multi-output pipeline'' can be partially evaluated with the following rule:
\[
\infer {\sepredone {\pipeM c l r} {\mval i {[b/l]r}}} {\reduce c {(i,b)}}
\]
We require that $c$ (the \emph{combined term}) reduce to the tuple $(i,b)$,
where $i$ becomes the immediate result of the masked value,
and where $b$ is plugged into the open variable of the resumer.
We refer to $b$ as the \emph{boundary value} of the split since it carries information between stage~\bbone\ and stage~\bbtwo\ subcomputations.

The splitting algorithm for stage~\bbone\ terms, as specified by the
judgment $e \splitonesym \pipeM c l r$ in \cref{fig:termSplit}, proceeds
recursively on the structure of~$e$.
Our goal for the transformation is that 
if $\diaone e {\xi,v}$ and $[\xi,v] \vsplito \mval i r$, 
then $\splitone e A {\pipeM c l {r'}}$ and $\sepredone {\pipeM c l {r'}} {\mval i {r''}}$,
where $r \equiv r''$ in some standard equivalence.
Intuitively, translation and evaluation should commute.

When $e$ is a terminal (a variable or base constant)
splitting yields a combined term formed by tupling $e$ with a \texttt{()} precomputation, and the trivial resumer \texttt{()}. (Stage~\bbone\ terminals, by definition, contain no stage~\bbtwo\ subcomputations.)  For example, the integer constant \texttt{3} splits into the combined term \texttt{(3,())} and resumer \texttt{\_=>()}.

For all non-terminals (except $\next$),
splitting descends into $e$, recursively splitting its $n$ subterms
to produce their respective combined terms $c_1,\ldots,c_n$ and resumers $r_1, \ldots, r_n$.
The combined term of $e$ is formed by binding $c_1,\ldots,c_n$
to the patterns $(y_1,z_1),\ldots,(y_n,z_n)$
to isolate \bbone-results from boundary values. Then,
the \bbone-result of $e$ is formed by replacing $e$'s subterms with $y_1,\ldots,y_n$.
The resumer binds the boundary values $b_1,\ldots,b_n$ to an
argument $(l_1,\ldots,l_n)$ in a term that has the same structure
of~$e$ but where each subterm is replaced by its resumer ($r_i$'s).

@@

\end{abstrsyn}
